---
title: "Bayesian Mortality Modeling"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bayesian Mortality Modeling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(dev = "png", dev.args = list(type = "cairo-png"))
```

```{r setup}
library(StanMoMo)
```



## Introduction: Bayesian Lee-Carter with Stan

First, we explain how to estimate the Lee-Carter model using the `lc_stan` function in the `StanMoMo` package.

For illustration, the package already includes the object `FRMaleData` containing deaths (`FRMaleData$Dxt`) and central exposures (`FRMaleData$Ext`) of French males for the period 1816-2017 and for ages 0-110 extracted from the [Human Mortality Database](https://www.mortality.org/). In this example, we focus on ages 50-90 and the period 1970-2017. This can be obtained via: 

```{r}
ages.fit<-50:90
years.fit<-1970:2017
deathFR<-FRMaleData$Dxt[formatC(ages.fit),formatC(years.fit)]
exposureFR<-FRMaleData$Ext[formatC(ages.fit),formatC(years.fit)]
```

As a reminder, the Lee-Carter model assumes that the log death rates are given by 
$$ 
\log \mu_{xt}=\alpha_x+\beta_x \kappa_t
$$
To ensure identifiability of the model, we assume that 
$$
\sum_x \beta_x=1,\kappa_1=0
$$

Moreover, we assume that the period parameter follows a first order autoregressive process (AR(1)) with linear trend:
$$ 
\kappa_t \sim \mathcal{N}(c+\delta t+\rho \kappa_{t-1},\sigma)
$$
We note that the random walk with drift appears as a special case. The choice of priors for each parameter can be found in the details of each function.

## Bayesian Estimation

All the models can be estimated either under a Poisson model with argument `family = "poisson"` or under a Negative-Binomial model which includes an additional overdispersion parameter $\phi$ with argument `family="nb"`:
$$
D_{xt} \sim Poisson (e_{xt}\mu_{xt}) \text{ or } D_{xt} \sim NB (e_{xt}\mu_{xt},\phi)
$$
Given the matrix of deaths `deathFR` and the matrix of central exposures `exposureFR`, the model can be estimated and forecast for the next 10 years under a Poisson model by a simple call to the `lc_stan` function: 

```{r}
fitLC=lc_stan(death = deathFR,exposure=exposureFR, forecast = 10, family = "poisson",cores=4)
```

By default, Stan samples four Markov chains with 2000 iterations with 1000 warmup iterations for each chain (hence, 4000 draws in total). The argument `cores=4` ensures that the four chains are sampled in parallel for time efficiency. The output is an object of class `stanfit` (cf. [the **rstan** package](http://mc-stan.org/rstan/)) which contains the posterior draws of each parameter, the posterior log likelihoods as well as the mortality forecasts.

The user can have access to an interface for interactive MCMC diagnostics, plots and tables helpful for analyzing posterior samples through the `shinystan` package.

```{r,eval=FALSE}
library(shinystan)
launch_shinystan(fitLC)
```

We can also plot the sampling for each parameter with the use of the `bayesplot` package. For instance, the posterior uncertainty intervals for $\alpha_x$, for $x=50, 51, 52, 53$, based on the 4000 draws can be obtained via 
```{r,warning=FALSE}
library(bayesplot)
posterior <- as.array(fitLC)
mcmc_areas(posterior,pars = c("a[1]", "a[2]", "a[3]", "a[4]"),prob = 0.8, prob_outer = 0.99, point_est = "mean")
```

We can also use the output to produce fan charts depicting the uncertainty around each parameter of the model fit.
```{r , echo=TRUE,warning=FALSE}
library("fanplot")
library("RColorBrewer")
library(latex2exp)
#Extract samples from the fitted model
params<-rstan::extract(fitLC)
#Alpha
plot(ages.fit, colMeans(params$a), ylim=range(params$a),ylab=TeX("$\\alpha_x$"), xlab="Age: x")
fan(data=params$a, start=ages.fit[1],type = "interval", ln=NULL,probs = c(0.5,0.8,0.95,0.995),
    fan.col = colorRampPalette(c("Orange", "white")))
#Beta
plot(ages.fit, colMeans(params$b), ylim=range(params$b),ylab=TeX("$\\beta_x$"), xlab="Age: x")
fan(data=params$b, start=ages.fit[1],type = "interval", ln = NULL,probs = c(0.5,0.8,0.95,0.995),
    fan.col = colorRampPalette(c("red", "white")))
#Kappa
plot(years.fit, colMeans(params$k), ylim=range(params$k),ylab=TeX("$\\kappa_t$"), xlab="Year: t")
fan(data=params$k, start=years.fit[1],type = "interval",ln = NULL, probs = c(0.5,0.8,0.95,0.995),
    fan.col = colorRampPalette(c("blue", "white")))
```

## Forecasting

Predictions of death rates for the next 10 years with confidence intervals can be obtained as follows: 

```{r , echo=TRUE}
# Resize the forecast deaths rates as an array "Number of draws X Ages X Years to predict"
samplesize<-4000
years.predict<-2018:2027
pred<-array(params$mufor,dim=list(samplesize,length(ages.fit),length(years.predict)),
            dimnames = list(c(1:samplesize),formatC(ages.fit),formatC(years.predict)))
#Fan plots for ages 65,75,85
probs = c(0.5,0.8,0.95,0.995)
qxt <- deathFR / exposureFR
matplot(years.fit, t(qxt[c("65", "75", "85"), ]),
        xlim = c(1970, 2027), ylim = c(0.0025, 0.2), pch = 20, col = "black",
        log = "y", xlab = "year", ylab = "death rate (log scale)")
fan(pred[,"65" , ], start = 2018, probs = probs, n.fan = 4,
    fan.col =  colorRampPalette(c("green", "white")), ln = NULL)
fan(pred[,"75" , ], start = 2018, probs = probs, n.fan = 4,
    fan.col =  colorRampPalette(c("red", "white")), ln = NULL)
fan(pred[,"85" , ], start = 2018, probs = probs, n.fan = 4,
    fan.col =  colorRampPalette(c("blue", "white")), ln = NULL)
text(1980, qxt[c("65", "75", "85"), "2000"],
     labels = c("x = 65", "x = 75", "x = 85"))
```

## Other models

In a similar fashion, the models RH, APC, CBD and M6 can be fitted with the following functions: 

| Model| Predictor | 
|------:|:-----|
|   LC  |  $\log \mu_{xt} = \alpha_x + \beta_x\kappa_t$ | 
|   RH  |  $\log \mu_{xt} = \alpha_x + \beta_x\kappa_t+\gamma_{t-x}$ | 
|   APC |  $\log \mu_{xt} = \alpha_x + \kappa_t +\gamma_{t-x}$ |   
|   CBD |  $\log \mu_{xt} = \kappa_t^{(1)} + (x-\bar{x})\kappa_t^{(2)}$ |   
|   M6  |  $\log \mu_{xt} = \kappa_t^{(1)} + (x-\bar{x})\kappa_t^{(2)}+\gamma_{t-x}$ |   

```{r,eval=FALSE}
fitRH=rh_stan(death = deathFR,exposure=exposureFR, forecast = 10, family = "poisson",cores=4)
fitAPC=apc_stan(death = deathFR,exposure=exposureFR, forecast = 10, family = "poisson",cores=4)
fitCBD=cbd_stan(death = deathFR,exposure=exposureFR, age=ages.fit, forecast=10,family = "poisson",cores=4)
fitM6=m6_stan(death = deathFR,exposure=exposureFR, age=ages.fit,forecast = 10, family = "poisson",cores=4)
```
