---
title: "Bayesian Mortality Modeling"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bayesian Mortality Modeling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(dev = "png", dev.args = list(type = "cairo-png"))
```

```{r setup}
library(StanMoMo)
```



## Introduction: Bayesian Lee-Carter with Stan

First, we explain how to estimate the Lee-Carter model using the `lc_stan` function in the `StanMoMo` package.

For illustration, the package already includes the object `FRMaleData` containing deaths (`FRMaleData$Dxt`) and central exposures (`FRMaleData$Ext`) of French males for the period 1816-2017 and for ages 0-110 extracted from the [Human Mortality Database](https://www.mortality.org/). In this example, we focus on ages 50-90 and the period 1970-2017. This can be obtained via: 

```{r}
ages.fit<-50:90
years.fit<-1970:2017
deathFR<-FRMaleData$Dxt[formatC(ages.fit),formatC(years.fit)]
exposureFR<-FRMaleData$Ext[formatC(ages.fit),formatC(years.fit)]
```

As a reminder, the Lee-Carter model assumes that the log death rates are given by 
$$ 
\log \mu_{xt}=\alpha_x+\beta_x \kappa_t
$$
To ensure identifiability of the model, we assume that 
$$
\sum_x \beta_x=1,\kappa_1=0
$$

Moreover, we assume that the period parameter follows a random walk with drift:
$$ 
\kappa_t \sim \mathcal{N}(c+ \kappa_{t-1},\sigma)
$$
The choice of priors for each parameter can be found in the documentation of each function.

## Bayesian Estimation

All the parameters can be estimated either under a Poisson model with argument `family = "poisson"` or under a Negative-Binomial model which includes an additional overdispersion parameter $\phi$ with argument `family="nb"`:
$$
D_{xt} \sim Poisson (e_{xt}\mu_{xt}) \text{ or } D_{xt} \sim NB (e_{xt}\mu_{xt},\phi)
$$
Given the matrix of deaths `deathFR` and the matrix of central exposures `exposureFR`, we can estimate the parameters and obtain death rates forecasts for the next 10 years under a Poisson model by a simple call to the `lc_stan` function: 

```{r}
fitLC=lc_stan(death = deathFR,exposure=exposureFR, forecast = 10, family = "poisson",cores=2)
```

By default, Stan samples four Markov chains of 2000 iterations. For each chain, there are 1000 warmup iterations (hence, 4000 post warm-up draws in total). One can use the argument `cores=4` to ensure that the four chains are sampled in parallel for time efficiency. The output is an object of class `stanfit` (cf. [the **rstan** package](https://mc-stan.org/rstan/articles/stanfit_objects.html)) which contains, among others, posterior draws, posterior summary statistics and convergence diagnostics. 

The easiest way to extract the posterior draws is to call the `rstan::extract` function which returns a list with named components corresponding to the model parameters.
```{r,warning=FALSE}
#Extract model parameters from the fitted model
params<-rstan::extract(fitLC)
print(names(params))
```

Among these parameters, we find

* `a` : $\alpha_x$.
* `b` : $\beta_x$.
* `k` : $\kappa_t$.
* `k_p` : forecasts of $\kappa_t$.
* `mufor` : forecasts of death rates $\mu_{xt}$.
* `log_lik` : log-likelihoods.
* `phi` : overdispersion parameter for the NB model. 
* `c` and `sigma` : drift and standard deviation of the random walk with drift.

The user can have access to an interface for interactive MCMC diagnostics, plots and tables helpful for analyzing posterior samples through the `shinystan` package (fore more details, you can check the [shinystan web page](http://mc-stan.org/shinystan/))

```{r,eval=FALSE}
library(shinystan)
launch_shinystan(fitLC)
```


As an illustration, we produce fan charts depicting the parameter uncertainty from the Lee-Carter model with the `fanplot` package.

```{r , echo=TRUE,warning=FALSE}
library("fanplot")
library("RColorBrewer")
library(latex2exp)
#Alpha
plot(ages.fit, colMeans(params$a), ylim=range(params$a),ylab=TeX("$\\alpha_x$"), xlab="Age: x")
fan(data=params$a, start=ages.fit[1],type = "interval", ln=NULL,probs = c(0.5,0.8,0.95,0.995),fan.col = colorRampPalette(c("Orange", "white")))
#Beta
plot(ages.fit, colMeans(params$b), ylim=range(params$b),ylab=TeX("$\\beta_x$"), xlab="Age: x")
fan(data=params$b, start=ages.fit[1],type = "interval", ln = NULL,probs = c(0.5,0.8,0.95,0.995),  fan.col = colorRampPalette(c("red", "white")))
#Kappa
plot(years.fit, colMeans(params$k), ylim=range(params$k),ylab=TeX("$\\kappa_t$"), xlab="Year: t")
fan(data=params$k, start=years.fit[1],type = "interval",ln = NULL, probs = c(0.5,0.8,0.95,0.995),fan.col = colorRampPalette(c("blue", "white")))
```

## Forecasting

All forecast death rates $\mu_{xt}$ are stored in the output `mufor`. We point out that the output `mufor` is a matrix of dimension $N$ $\times$ $(A . M)$ where  the number of rows $N$ is the posterior sample size and the number of columns $A.M$ is the product of $A$ the age dimension and $M$ the number of forecast years.

Predictions of death rates for the next 10 years with prediction intervals can be obtained as follows: 

```{r , echo=TRUE}
# Resize the forecast deaths rates as an array "Number of draws X Ages X Years to predict"
samplesize<-4000
years.predict<-2018:2027
pred<-array(params$mufor,dim=list(samplesize,length(ages.fit),length(years.predict)),dimnames = list(c(1:samplesize),formatC(ages.fit),formatC(years.predict)))
#Fan plots for ages 65,75,85
probs = c(0.5,0.8,0.95,0.995)
qxt <- deathFR / exposureFR
matplot(years.fit, t(qxt[c("65", "75", "85"), ]),
        xlim = c(1970, 2027), ylim = c(0.0025, 0.2), pch = 20, col = "black",
        log = "y", xlab = "year", ylab = "death rate (log scale)")
fan(pred[,"65" , ], start = 2018, probs = probs, n.fan = 5,
    fan.col =  colorRampPalette(c("green", "white")), ln = NULL)
fan(pred[,"75" , ], start = 2018, probs = probs, n.fan = 5,
    fan.col =  colorRampPalette(c("red", "white")), ln = NULL)
fan(pred[,"85" , ], start = 2018, probs = probs, n.fan = 5,
    fan.col =  colorRampPalette(c("blue", "white")), ln = NULL)
text(1980, qxt[c("65", "75", "85"), "2000"],
     labels = c("x = 65", "x = 75", "x = 85"))
```

## Other models

In a similar fashion, the models RH, APC, CBD and M6 can be estimated with the following functions: 

| Model| Predictor | 
|:------|:-----|
|   LC  |  $\log \mu_{xt} = \alpha_x + \beta_x\kappa_t$ | 
|   RH  |  $\log \mu_{xt} = \alpha_x + \beta_x\kappa_t+\gamma_{t-x}$ | 
|   APC |  $\log \mu_{xt} = \alpha_x + \kappa_t +\gamma_{t-x}$ |   
|   CBD |  $\log \mu_{xt} = \kappa_t^{(1)} + (x-\bar{x})\kappa_t^{(2)}$ |   
|   M6  |  $\log \mu_{xt} = \kappa_t^{(1)} + (x-\bar{x})\kappa_t^{(2)}+\gamma_{t-x}$ |   

```{r,eval=FALSE}
fitRH=rh_stan(death = deathFR,exposure=exposureFR, forecast = 10, family = "poisson",cores=4)
fitAPC=apc_stan(death = deathFR,exposure=exposureFR, forecast = 10, family = "poisson",cores=4)
fitCBD=cbd_stan(death = deathFR,exposure=exposureFR, age=ages.fit, forecast=10,family = "poisson",cores=4)
fitM6=m6_stan(death = deathFR,exposure=exposureFR, age=ages.fit,forecast = 10, family = "poisson",cores=4)
```

For information, the estimated parameters $\kappa_t^{(1)}$, $\kappa_t^{(2)}$ and $\gamma_{t-x}$ are respectively stored in the variables `k`,`k2` and `g`. The forecast parameters are respectively stored in the variables `k_p`,`k2_p` and `g_p`.

You can check the next vignette to know more about model averaging techniques based on leave-future-out validation.

