---
title: "StanMoMo"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The `StanMoMo` package performs Bayesian **Mo**rtality **Mo**deling with **Stan**, which is a C++ package for performing full Bayesian inference (see <https://mc-stan.org/>). The current package supports a variety of popular mortality models: the Lee-Carter (LC) model, the Renshaw-Haberman model (LC with cohort effect), the Age-Period-Cohort (APC) model, the Cairns-Blake-Dowd (CBD) model and the M6 model (CBD with cohort effect). By a simple function call, the user obtains the MCMC simulations for each parameter, the log likelihoods and death rates predictions. Moreover, the package includes tools for model selection and Bayesian model averaging by future-out cross-validation.

## Installation

To install `StanMoMo` from GitHub you will need `devtools`:

``` r
install.packages("devtools")
devtools::install_github('kabarigou/StanMoMo')
```

After installing the package, you have to load the package via: 
```{r}
library(StanMoMo)
```

## Important note

The main purpose of this package is to provide users high-level functions for estimating and forecasting mortality models in a Bayesian setting without requiring any knowledge of the Stan modeling language. This package depends on the [rstan](https://mc-stan.org/rstan/) package, which translates the Stan model to C++ code, which is compiled into a dynamic shared object (DSO). During the package installation, the mortality models are pre-compiled and therefore you need a C++ compiler on your machine (for instance, Rtools for Windows or Xcode on Mac, see [here](https://github.com/stan-dev/rstan/wiki/RStan-Getting-Started) for more details). Once the package will be released on CRAN (date not defined yet), all mortality models will be already pre-compiled in order to run immediately when called, avoiding compilation time.

If you have any comments or suggestions about the package, feel free to email <karim.barigou@univ-lyon1.fr>






